---
title: "Use geotagged jpeg images as map backgrounds"
output: 
  html_document: 
    smart: no
    self_contained: no
    number_sections: no
    toc: no
---

<style type="text/css">
  body{
  font-size: 11pt;
}
</style>

# Introduction

Sometimes we can obtain the best map background images in the form of
georeferenced images, such as geoTIFF or jpeg. In this example we'll use a jpeg
file downloaded from the excellent
[NearMap](https://www.nearmap.com/au/en){target="_blank"} site (my university
has a subscription 😊). Specialist mapping websites may have better resolution,
or have aerial photographs from the most appropriate time of year for our
mapping exercise.

First we load the packages that we need:

```{r load-packages, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(rgdal)
library(viridis)
library(stars)
library(prettymapr)
```

Let's just have a look at the image file using the `GDALinfo` function from the
`rgdal` package:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
GDALinfo("EPSG32750_Date20230226_Lat-31.918263_Lon115.944872_Mpp0.597.jpg") # just file info
```

We notice that the jpeg file itself doesn't have spatial information, even 
though we may have downloaded it as a georeferenced image. With jpeg images, 
georeferencing comes in the form of a companion ('sidecar') file, with the same
name but having the `.jgw` extension. We can read this directly, but the
functions we will use below will look for this file and use the spatial
information directly. The `.jgw` file has 6 lines:

1. length of a pixel in the x direction (horizontal)
2. angle of rotation (is usually 0 or ignored)
3. angle of rotation (is usually 0 or ignored)
4. negative length of a pixel in the y direction (vertical)
5. x coordinate at centre of pixel in the top left corner of the image
6. y coordinate at centre of pixel in the top left corner of the image

So long as the file names of the `.jpg` and `.jpw` files match, the raster
extent and coordinates are obtained or calculated from the values in the `.jpw`
file.

# Reading the raster spatial information in the georeferenced jpeg image

We use the `stars` package, which plays nicely with the `sf` package, to do most
of the work (the `sf` package was introduced on [another page on maps in
**R**](Rmaps.html). The `read_stars()` function will find and use the
information in the `.jpw` file without us specifying that file explicitly.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
(AF_nearmap <- 
  read_stars("EPSG32750_Date20230226_Lat-31.918263_Lon115.944872_Mpp0.597.jpg"))
```

We notice that, apart from the long and informative filename, the resulting 
`stars` object does not have a coordinate reference system included, so we need 
to add the appropriate one manually:

```{r}
st_crs(AF_nearmap) <- st_crs(32750)
strtrim(st_crs(AF_nearmap),30)
```

The messy output above shows that we have what we expect: UTM Zone 50 (south),
based on the [WGS84 datum](https://support.virtual-surveyor.com/en/support/solutions/articles/1000261351-what-is-wgs84-){target="_blank"}.

From the summary of the `read_stars()` output above (*i.e*. `band    1    3`), 
we can see that the image has three bands of information, corresponding to the
red, green and blue colour channels in the jpeg image.

To plot as a combined colour image we need to merge these 3 channels into in 
single data layer of colour information; fortunately, there is a function 
for this in the `stars` package: `st_rgb()`:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
AF_rgb <- st_rgb(AF_nearmap)
print(AF_rgb)
```

We need some data to plot, so let's read some

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
git <- "https://github.com/Ratey-AtUWA/cybloRg/raw/main/"
afs1922 <- read.csv(paste0(git,"afs1922edit.csv"), stringsAsFactors = TRUE)
afs1922$Year <- as.factor(afs1922$Year)
(afsREE <- st_as_sf(x = afs1922[,c("Year","Easting", "Northing","Al","Fe","REE")],
         coords = c("Easting", "Northing"), crs = st_crs(32750)))
```

Now we can plot a map with some data on it! We use the `image()` function to
plot the stars raster; we could also use `plot()` if we specify the argument
`reset = FALSE`.

```{r afr-nearmap, fig.align='center', fig.height=6.6, fig.width=10, message=FALSE, warning=FALSE, fig.cap="Figure 1: Map of sampling locations at Ashfield Flats Reserve from 2019-2022 plotted on a Nearmap spatial raster.", results='hold'}
palette(c("white",viridis::viridis(8), "black","transparent"))
par(oma=c(3,3,1,1), mar=c(0,0,0,0), mgp=c(1.7,0.3,0), tcl=0.25,
    lend="square", ljoin="mitre")
image(AF_rgb, main="", axes=TRUE)
axis(3,labels=F); axis(4, labels=F)#; grid(lty=3, col=1)
mtext("Easting (UTM Zone 50S, m)",1,1.7,font=2,cex=1.2)
points(st_coordinates(afsREE), pch=c(21:24)[afsREE$Year],
       bg=c(3,5,7,8)[afsREE$Year], col=1, cex=c(1.2,1,1,1)[afsREE$Year], lwd=2)
addnortharrow(pos="topright",padin = c(0.62,0.62), 
              text.col = 10, border = 10, scale=1.4)
addnortharrow(pos="topright",padin = c(0.6,0.6), 
              text.col = 1, border = 1, scale=1.4)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 11,label.col = 10,
            padin = c(0.62,0.62),htin = 0.2, label.cex = 1.5, bar.cols=c(11,11),
            widthhint = 0.15)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 1,label.col = 1,
            padin = c(0.6,0.6),htin = 0.2, label.cex = 1.5, widthhint = 0.15)
legend("topleft", title=expression(bold("Ashfield Flats Reserve")),
       legend=c("Map source: NearMap Feb 2023",
                "Projection: UTM Zone 50 S     ",
                "EPSG: 32750",
                "Geoid: WGS84"),
       box.col = "white",inset=c(0.03,0.04), cex=1, y.intersp = 1)
legend("bottomright", title=expression(bold("Year")),
       legend=levels(afsREE$Year), text.col = 1, pch=c(21,22,23,24), 
       pt.bg=c(3,5,7,8), col=1, box.col=1, pt.lwd=2, pt.cex=c(1.2,1,1,1), 
       bg = "#808070",inset=c(0.08,0.12), cex=1, y.intersp = 1)
```

<p>&nbsp;</p>

For comparison we can also use the `add.geom =` argument in the `image()` or
`plot()` function as an alternative way to plot the points.

```{r afr-nearmap-addgeom, fig.align='center', fig.height=6.6, fig.width=10, message=FALSE, warning=FALSE, fig.cap="Figure 2: Map of sampling locations at Ashfield Flats Reserve from 2019-2022 plotted on a Nearmap spatial raster using `plot.geom` as an argument to `image()`.", results='hold'}
palette(c("white",viridis::viridis(12, direction=-1)[1:8], "black","transparent"))
par(oma=c(3,3,1,1), mar=c(0,0,0,0), mgp=c(1.7,0.3,0), tcl=0.25,
    lend="square", ljoin="mitre")
image(AF_rgb, main="", axes=TRUE, add.geom=afsREE)
axis(3,labels=F); axis(4, labels=F)#; grid(lty=3, col=1)
mtext("Easting (UTM Zone 50S, m)",1,1.7,font=2,cex=1.2)
addnortharrow(pos="topright",padin = c(0.62,0.62), 
              text.col = 10, border = 10, scale=1.4)
addnortharrow(pos="topright",padin = c(0.6,0.6), 
              text.col = 1, border = 1, scale=1.4)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 11,label.col = 10,
            padin = c(0.62,0.62),htin = 0.2, label.cex = 1.5, bar.cols=c(11,11),
            widthhint = 0.15)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 1,label.col = 1,
            padin = c(0.6,0.6),htin = 0.2, label.cex = 1.5, widthhint = 0.15)
legend("topleft", title=expression(bold("Ashfield Flats Reserve")),
       legend=c("Map source: NearMap Feb 2023",
                "Projection: UTM Zone 50 S     ",
                "EPSG: 32750",
                "Geoid: WGS84"),
       box.col = "white",inset=c(0.03,0.04), cex=1, y.intersp = 1)
```

<p>&nbsp;</p>

The result is not as satisfying as adding points separately using `points()`. We
could also add our data in a way that shows values of a variable, as described 
in the [page on maps in **R**](Rmaps.html).

## Packages

Bivand R, Keitt T, Rowlingson B (2022). *rgdal: Bindings for the 'Geospatial' Data Abstraction*
  *Library*. R package version 1.6-2, [https://CRAN.R-project.org/package=rgdal](https://CRAN.R-project.org/package=rgdal){target="_blank"}.
  
Dunnington D (2022). _prettymapr: Scale Bar, North Arrow, and Pretty Margins in
R_. R package version 0.2.4,
[https://CRAN.R-project.org/package=prettymapr](https://CRAN.R-project.org/package=prettymapr){target="_blank"}.
  
Garnier S., Ross N., Rudis R., Camargo A.P., Sciaini M., and Scherer C.
  (2021). *Rvision - Colorblind-Friendly Color Maps for R*. R package version 0.6.2. [https://sjmgarnier.github.io/viridis/](https://sjmgarnier.github.io/viridis/){target="_blank"}

Hijmans R (2022). _raster: Geographic Data Analysis and Modeling_. R package
version 3.6-11,
[https://CRAN.R-project.org/package=raster](https://CRAN.R-project.org/package=raster){target="_blank"}.

Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data. *The R*
  *Journal* **10** (1), 439-446, [https://doi.org/10.32614/RJ-2018-009](https://doi.org/10.32614/RJ-2018-009){target="_blank"}

Pebesma E (2022). _stars: Spatiotemporal Arrays, Raster and Vector Data Cubes_.
R package version 0.6-0,
[https://CRAN.R-project.org/package=stars](https://CRAN.R-project.org/package=stars){target="_blank"}.
