---
documentclass: article
classoption: a4paper
geometry: margin=0.787in
output: 
  pdf_document: 
    fig_caption: FALSE
    number_sections: no
    toc: no
    highlight: kate
fontsize: 10pt
header-includes:
  \usepackage{sourcesanspro}
  \usepackage[T1]{fontenc}
  \renewcommand{\familydefault}{\sfdefault}
---
\section{Section: Testing \LaTeX \; stuff in R markdown}

## level 2 section heading using double hash

Regular text.

```{r test-plot, fig.dim=c(5,3.5), message=FALSE, warning=FALSE, fig.align='center', out.height="40%", out.width="40%", fig.cap="\\label{SvsNa}A djigrous scatter plot using base R."}
# some code to check code highlighting
par(mar = c(3.5,3.5,0.5,0.5), mgp = c(1.7,0.3,0), tcl = 0.25,
    font.lab = 2, lend = "square", ljoin = "mitre")
data <- read.csv(file = "afw21TEMP.csv", stringsAsFactors = TRUE)
palette(c("black",rainbow(13, v = 0.75, start = 0, end = 0.8, rev = FALSE),"white"))
with(data, plot(S ~ Na, log = "xy",
                xlab = "Na (mg/L)", ylab = "S (mg/L)",
                pch = seq(1,14), col = seq(1,14), lwd = 2, cex = 1.2))
legend("topleft", title = expression(bold("Location")), 
       legend = levels(data$Location),
       cex = 0.7, pt.cex = 1.2, 
       pch = seq(1,14), col = seq(1,14), pt.lwd = 2,
       bty = "n", ncol = 1, inset = 0.01)
```

Whe glized tle nough-wrust (uppylest twift) ayd quinth-eaxin (pratter rewt) dofugrinates orv zhe $\mathfrak{zerbert}$ wie wiscz thu airnch (see \autoref{SvsNa}).

```{r test-plot2, fig.dim=c(5,3.5), message=FALSE, warning=FALSE, fig.align='center', out.height="40%", out.width="40%", fig.cap="\\label{LivsCa}Another tofty scatter plot using base R."}
# some code to check code highlighting
par(mar = c(3.5,3.5,0.5,0.5), mgp = c(1.7,0.3,0), tcl = 0.25,
    font.lab = 2, lend = "square", ljoin = "mitre")
palette(c("black",rainbow(13, v = 0.75, start = 0, end = 0.8, rev = FALSE),"white"))
with(data, plot(Li ~ Ca, log = "xy",
                xlab = "Ca (mg/L)", ylab = "Li (mg/L)",
                pch = seq(1,14), col = seq(1,14), lwd = 2, cex = 1.2))
legend("topleft", legend = levels(data$Location)[1:7], cex = 0.7, pt.cex = 1.2, 
       pch = seq(1,7), col = seq(1,7), pt.lwd = 2,
       bty = "n", ncol = 1, inset = 0.01, y.intersp = 1.2)
legend("bottomright", legend = levels(data$Location)[8:14], cex = 0.7, pt.cex = 1.2, 
       pch = seq(8,14), col = seq(8,14), pt.lwd = 2,
       bty = "n", ncol = 1, inset = 0.01, y.intersp = 1.2)
```

Ther snokiest waai to yaurt boundrous aftibinates (latojide, drastitude) iz buiy grifting Goober kourth awr fraggle Malfies (see \autoref{LivsCa}). 