---
title: "Introduction to R for ENVT4461"
output:
  bookdown::html_document2:
    fig_width: 6
    fig_height: 4
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r addimg function, include=FALSE}
library(png)
addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

```{r save-default-par, include=FALSE}
defpar <- par(no.readonly = TRUE)
```

```{r page 1 header hide code, fig.height=1.5, fig.width=10, echo=FALSE, out.width="100%", fig.align='right', results='hold'}
logo <- readPNG("UWA logo_text_V_wsL.png")
par(mar = c(0,0,0,0))
layout(matrix(c(1,1,1,1,2),nrow = 1))

plot(c(0,1),c(0,1), axes=F, type="n",xaxt="n", yaxt="n",ann=F)
text(-0.025,0.9, pos = 4, cex = 2.6, font = 2, col = "#003087", 
     labels="School of Agriculture and Environment")
text(-0.025,0.6, pos = 4, cex = 2.2, col = "#997500", 
     labels="R for data analysis and visualization in Environmental Science")
# text(-0.025,0.4, pos = 4, font = 3, cex = 1.4, col = 12,
#      labels="Geochemical data for international samples of whole rocks (USGS)")
text(-0.025,0.1, pos = 4, font = 3, family = 'serif', cex = 1.5, col = "#636363",
     labels="James Fogarty and Andrew Rate, School of Agriculture and Environment")
plot(1,1, axes=F, type="n",xaxt="n", yaxt="n",ann=F)
addImg(logo, x = 1.2, y = 1, width = 0.5)
par(mar = c(3.5,3.5,0.5,0.5))
layout(matrix(c(1),nrow = 1))
```

# computer lab files

<p>&nbsp;</p>

<img style="height: auto;" width="136" height="200" alt="brushtail possum" align="right" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42743113_1/xid-42743113_1" /><span style="font-size: 12pt;">The **Distributions chapter in the Textbook**, also available here:&nbsp;<a rel="noopener" target="_blank" href="https://saestatsteaching.tech/section-distributions">https://saestatsteaching.tech/section-distributions</a>&nbsp;covers the theory elements that underlie the worked script example material</span>

- &nbsp;<a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747553_1/xid-27747553_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42743115_1/xid-42743115_1" alt="File" />&nbsp;Computer lab 2 exercise 1 distributions.R</a>

- <a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747554_1/xid-27747554_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42743115_1/xid-42743115_1" alt="File" />&nbsp;Computer lab 2 exercise 2 histogram.R</a>

- &nbsp;<a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747555_1/xid-27747555_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42743115_1/xid-42743115_1" alt="File" />&nbsp;Computer lab 2 exercise 3 data transformations.R</a>

- <a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747580_1/xid-27747580_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42740840_1/xid-42740840_1" alt="File" />&nbsp;mine.sample.csv</a>

- <a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747587_1/xid-27747587_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42740840_1/xid-42740840_1" alt="File" />&nbsp;sv2017_original.csv</a>

- <a rel="noopener" href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-27747592_1/xid-27747592_1" target="_blank"><img style="height: auto;" width="32" height="32" src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42740840_1/xid-42740840_1" alt="File" />&nbsp;wingspanMALES.csv</a>

<p>&nbsp;</p>

## Videos and other learning materials

<table style="width: 100%;" border="0">
<tbody>
<tr>
<td valign="top" style="width: 56%;">
<div style="margin-left: 30px; font-size: 11px; margin-bottom: 5px;"><iframe width="480" height="360" src="https://echo360.net.au/media/fcbcb5dd-ac16-4c94-910c-4ca1b3f3d499/public?autoplay=false&amp;automute=false" title="E3361+cat+and+transformations.mp4" allowfullscreen="allowfullscreen" webkitallowfullscreen="webkitallowfullscreen" mozallowfullscreen="mozallowfullscreen"></iframe>&nbsp; <br />The 'cat video' version ( *and thanks to [Dream Theater](http://dreamtheater.net/){target="_blank"}</a> for the mood music*)</div>
<div style="margin-left: 30px; font-size: 11px; margin-bottom: 5px;"><span style="background-color: #ffff80;">**<span style="font-size: 12pt; background-color: #ffff80;">Ignore the reference near the end of the video to a Short Report!!</span>**</span></div>
<p>&nbsp;</p>
</td>
<td valign="top" style="width: 44%; vertical-align: top;">

<span style="font-size: 12pt;"><a href="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-42740883_1/xid-42740883_1" target="_blank" rel="noopener"><img src="https://lms.uwa.edu.au/bbcswebdav/pid-3355657-dt-content-rid-40243195_1/xid-40243195_1" width="48" height="48" alt="pdf file icon" style="float: left;" /><span style="background-color: #ffff40;">**<span style="font-size: 10pt; background-color: #ffff40;">NEW </span>**</span>Distributions and transformations of variables - Annotated R code</a></span></p>

<span style="font-size: 12pt;"><span class="mce-nbsp-wrap" contenteditable="false">&nbsp;</span></span></p>

<span style="font-size: 12pt;">See Chapter 10&nbsp;in Riemann *et al*. (2008)</span><br /><span style="font-size: 12pt;">&ldquo;<a target="_blank" href="http://onesearch.library.uwa.edu.au/UWA:Blended:UWA_ALMA51321155750002101" rel="noopener">**Improving Data Behaviour for Statistical Analysis: Ranking and Transformations**</a>&rdquo;</span></p>
</td>
</tr>
</tbody>
</table>
