---
documentclass: article
classoption: a4paper
geometry: margin=0.7874in
output: 
  pdf_document: 
    fig_caption: yes
---

```{r load data & packages, message=FALSE, warning=FALSE, include=FALSE}
library(RcmdrMisc)
library(knitr)
library(Rmisc)
metal <- read.csv("https://raw.githubusercontent.com/Ratey-AtUWA/learningR/main/metal.csv", stringsAsFactors = TRUE)
metal$Title <- as.character(metal$Title)
```

## One page short report format: the report needs a title . . . . . don't forget your name and UWA ID . . . . .

**Introduction**. The objective was to analyse *xxx* data by applying *yyy* statistical / numerical procedure or creating *zzz* model supported by *jjj* checking, and then to interpret the results in terms of real biological-chemical-physical processes. At most 1-2 sentences; you can save space using 10pt font (like this).

**Methods**. A concise description of checking variable distributions and any transformations, and some details e.g. which tests for means comparisons and why, selection of predictors for models and why, and so on. You might need one or at most two sentences here as well, depending on the complexity of the methods and the nature of the report; you can save space by using single spacing (like this).

**Results**. Present mainly tables and figures with supporting text which refers to the correct figure/table. Some examples are below. Please include informative  captions for both Figures and Tables.

```{r make table of means and SE}
numSummary(metal[,c("Growl","Chug","Shred","Riffage")],
                    statistics=c("mean","se(mean"),groups=metal$Band)
```

```{r check equality of variance}
with(metal, bartlett.test(Growl, Band))
with(metal, bartlett.test(Chug, Band))
with(metal, bartlett.test(Shred, Band))
with(metal, bartlett.test(Riffage, Band))
```

```{r conduct oneway tests}
with(metal, oneway.test(Growl ~ Band, var.equal = FALSE))
with(metal, oneway.test(Chug ~ Band, var.equal = FALSE))
with(metal, oneway.test(Shred ~ Band, var.equal = FALSE))
with(metal, oneway.test(Riffage ~ Band, var.equal = FALSE))
```

```{r pairwise comparisons}
with(metal, pairwise.t.test(Growl, Band, pool.sd = FALSE))
with(metal, pairwise.t.test(Chug, Band, pool.sd = FALSE))
with(metal, pairwise.t.test(Shred, Band, pool.sd = FALSE))
with(metal, pairwise.t.test(Riffage, Band, pool.sd = FALSE))
```

```{r Chug boxplot with added info, fig.width=10, fig.height=5, fig.cap="Example of (left) box plot and (right) strip chart showing differences in chug score for selected bands. Try to make all the text in figures large enough to read!"}
par(mfrow=c(1,2), mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.25, font.lab = 2, 
    lend = "square", ljoin = "mitre", xpd = TRUE)
palette(c("black","red3","darkgreen","blue2","#DDEE99",
          "#FFCC99","#e0e0e0","white","grey40","thistle", "purple"))
require(Rmisc) 
ci0 <- group.CI(Chug~Band, data=metal)
boxplot(metal$Chug~metal$Band, varwidth=T, col=c(6,7,5),
        xlab = "Metal Band", ylab="Chug score",
        cex.lab=1.2, xaxt = "n") 
axis(1, tcl = 0.4, at = seq(1,3), 
     labels = as.character(levels(metal$Band)))
arrows(x0=seq(1,NROW(ci0)), y0=ci0[,3], y1=ci0[,2], 
       col=8, angle=90, length=0.1, lwd=6) # optional
arrows(x0=seq(1,NROW(ci0)), y0=ci0[,3], y1=ci0[,4], 
       col=8, angle=90, length=0.1, lwd=6) # optional
lines(seq(1,NROW(ci0)), ci0[,3], col=8, lwd=3, 
      type="c") # optional
lines(seq(1,NROW(ci0)), ci0[,3], col=4, lwd=1, 
      type="c", lty=3) # optional
arrows(x0=seq(1,NROW(ci0)), y0=ci0[,3], y1=ci0[,2], 
       col=4, angle=90, length=0.1, lwd=2)
arrows(x0=seq(1,NROW(ci0)), y0=ci0[,3], y1=ci0[,4], 
       col=4, angle=90, length=0.1, lwd=2)
points(seq(1,NROW(ci0)), ci0[,3], col=8, pch=16, lwd=2, cex=1.6)
points(seq(1,NROW(ci0)), ci0[,3], col=4, pch=16, lwd=2, 
       cex=1.2)
legend("topright", legend=c("Mean \u00B1 95% CI","Potential outliers"),
       pch=c(16,1), cex=1., pt.cex=c(1.4,1.2), col=c(4,1), lwd=c(1,NA),
       lty=c(3,NA), bty="n", inset=0.01, y.intersp=1.5)
rm(ci0)

with(metal, stripchart(Chug ~ Band, vertical = T, method = "jitter",
                       xlab = "Metal Band", ylab="Chug score", 
                       pch = c(0,1,2), col = c(1,2,4)))
```

```{r scatterplot matrix, fig.height=6, fig.width=6}
par(mfrow=c(1,1), mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.25, font.lab = 2, 
    lend = "square", ljoin = "mitre", xpd = TRUE, cex = 0.85)
palette(c("black","red3","darkgreen","blue2","#DDEE99",
          "#FFCC99","#e0e0e0","white","grey40","thistle", "purple"))
scatterplotMatrix(~ Growl + Chug + Shred + Riffage | Band, data = metal,
                  smooth = FALSE, col = c(1,2,4), pch = c(0,1,2))
```

```{r}
require(RcmdrMisc)
rcorr.adjust(metal[,c("Growl","Chug","Shred","Riffage")])
```

Correlations between Growl-Chug, Growl-Riffage, and Chug-Riffage were significant at p < 0.05 (Holm's adjusted values for multiple comparisons). There is a little more space here for some text describing your results if you need it. There is a little more space here for some text describing your results if you need it.

**Discussion**. The tendency for Gojira tracks to have greater metalness scores (Table 1, Figure 1) was expected, given their reported history in Death Metal (Balas, 2017). The exception was Shred (Table 1), where Megadeth and Triosphere both had significantly greater scores, consistent with the known virtuosity of the principal guitar players (Sonic Cathedral, 2010; Florino, 2017). This different trend in means for Chug was supported by the absence of any Chug-Shred correlation (Table 2), with only predominantly bass-register metalness (Growl, Chug, and Riffage) being significantly correlated.

### References (which can be in slightly smaller font)

Balas, J. 2017. \textit{GOJIRA - From an Underground Beast to the Very Top of the Metal World}.  https://www.hardwiredmagazine.com/gojira-from-the-beast-from-the-underground-to-the-very-top-of-the-metal-world/ 

Sonic Cathedral, 2010. \textit{Triosphere -- CD Review -- A Road Less Travelled}. https://www.soniccathedral.com/zine/index.php/cd-reviews/908-Triosphere%20-%20The%20Road%20Less%20Travelled. 

Florino, R. 2017. \textit{Think Piece: Why Megadeth's Dystopia Is So Important}. https://allaxess.com/think-piece-megadeths-dystopia-important/
