---
title: "Workshop 3"
subtitle: "Making and describing data frames"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Scope from LMS Class Schedule**

#### Learning goals

* Making data in R: vectors & data frames
* Adding new data frame columns
* Generating basic descriptive statistics
* Summarising variables in different ways

## Making data in R

In the previous session we covered creating of vector objects and data 
frames using R code, so here we will do some quick revision.

### Vectors

We've seen how to make vectors using the `c()` function to combine
vector elements; below we make a vector from a sequence using the
`seq()` function:

```{r make-vector-using-seq}
# make a vector called x1
x1 <- seq(from=10, to=200, by=10)
x1
```

We could abbreviate `x1 <- seq(from=10, to=200, by=10)` to 
`x1 <- seq(10, 200, 10)`, since `from`, `to`, and `by` are the first 
3&nbsp;options in the `seq()` function, so this is what R expects.

You could also try making a vector using the `rep()` (replicate)
function (run `?rep` in the console for help).

### Data frames

We made a data frame `brightstars` in the previous sessions using the 
`data.frame()` function (review that if you need to). Here's another 
example:

```{r make-data-frame-using-cbind, paged.print=FALSE}
x2 <- rep(c("a", "b", "c", "d"), 5)
x3 <- rep(c(TRUE, FALSE), each=10) # note the difference!
new_data <- data.frame(x1, x2, x3)
print(new_data)
```

Instead of defining the contents of the data frame in the
`data.frame()` function itself, we have used vectors of equal length
that already exist (`x1`, `x2`, and `x3`).

We can also name the data frame columns if desired, e,g.: 

```{r eval=FALSE}
new_data <- data.frame(Numbers=x1, Letters=x2, Logical=x3)
```

**OR** if we already have made the data frame:

```{r eval=FALSE}
colnames(new_data) <- c("Numbers", "Letters", "Logical")
```

If we wanted column 2 `x2` to be a factor, we would include the option 
`stringsAsFactors = TRUE`:

```{r make-data-frame-with-factor}
new_data <- data.frame(x1, x2, x3, stringsAsFactors = TRUE)
str(new_data)
```

### Adding new columns to a data frame

We often add columns to a data frame to calculate a new column from 
existing data. For example, we may want to log-transform a variable to 
make its distribution more symmetrical:

```{r add-log-column}
new_data$x1_log <- log10(new_data$x1) # note we prefer log to base 10
new_data$x1_log
cat("\n") # insert line break to separate output
str(new_data)
```

Note that there are equivalent ways to do this, as is often the case 
in R!

```{r add-log-a-different-way}
new_data$x1_log <- log10(new_data[,1])
new_data$x1_log
```

```{r add-log-another-way}
new_data$x1_log <- with(new_data, log10(x1))
new_data$x1_log
```

We can also add new column(s) from an existing object (the added
object must have the same number of values as each data frame column):

```{r add-colums-using-cbind}
x4 <- rep(c("V","W","X","Y","Z"), each=4)
new_data <- cbind(new_data, x4)
str(new_data)
```

Finally we can remove unwanted columns from a data frame by assigning
the special value `NULL` to them. For example, to return to our
initial data frame:

```{r remove-columns}
new_data[,4:5] <- NULL
str(new_data)
```

## Calculating basic statistics

### Averages and spreads

We're often interested in the distribution of a numeric variable,
expressed by the central "average" value and the spread of values
around the average. Average values are commonly the *mean* or
*median*, and the spread of values is often given as the
*standard deviation* or *variance*. We also get useful information on
the spread of values from the *minimum* and *maximum*.

First let's read in some data from a previous session:

```{r read-cities-data, paged.print=FALSE}
cities <- read.csv("cities.csv", stringsAsFactors = TRUE)
cities$City <- as.character(cities$City)
str(cities)
```

```{r mean}
mean(cities$Compact)
```

```{r median}
median(cities$Compact)
```

```{r standard-deviation}
sd(cities$Compact) # standard deviation
```

```{r variance}
var(cities$Compact) # variance
```

```{r minimum}
min(cities$Compact) # minimum
```

```{r maximum}
max(cities$Compact) # maximum
```

```{r range}
range(cities$Compact) # maximum AND minimum
```

Some of this information is calculated by using `summary()` for a 
single variable (i.e. column in a data frame):

```{r summary-numeric-var}
sum0 <- summary(cities$Compact)
sum0
```

We can access individual statistics this way, but it's easier to use
the individual functions.

```{r summary-extract-values}
sum0["Mean"]
```

> 💡Try calculating some basic statistics for other numeric variables 
> in the `cities` data frame...

**NOTE**: Some of these simple statistical functions will not handle 
missing values by default. In R the missing value code is `NA`. 
For example:

```{r mean-with-missing-value}
x5 <- c(1,2,4,8,16,NA,32,64,128)
mean(x5)
cat("\nNow add the na.rm=TRUE option\n")
mean(x5, na.rm = TRUE)
```

> 💡Try calculating the `median`, `sd`, `min`, `max`, 
> and `range` for `x5`!


### Other useful functions

```{r sum-numeric-columns}
sum(cities$Open)
sum(x5, na.rm=TRUE)
```

We can find missing (`NA`) values, or non-missing:

```{r find-missing}
is.na(x5)
```

```{r find-non-missing}
!is.na(x5)
```

If we want to know the row index number of any missing observation(s), 
we use `which()`:

```{r using-which}
which(is.na(x5))
cat("\n")
which(!is.na(x5))
```

### Using `which()` to count values matching a condition

This is an R method we will make use of to make tables for reports 
later. We just used `which()` above to find the rows which did not 
contain a missing value `NA` for a variable. In this case the 
*condition* was `!is.na()`, but we can impose any condition that makes 
sense for our data. Here are some examples:

```{r which-over-threshold}
c500 <- which(cities$Compact > 500)
c500
```

The next example uses the function `substr()` (*substring*) to shorten
city names to just the first letter:

```{r which-initial-letter}
# which city names begin with a selected letter?
let1 <- "N"
first <-which(substr(cities$City,1,1)==let1) 
first
```

We can use the object containing the row numbers in an index to view
the rows matching the condition (we also introduce another very useful
function, `length()`). See if you can follow what is happening here:

```{r which-in-index, paged.print=FALSE}
len1 <- length(c500)
cities[c500,1:5]
cat("There are",len1,"cities with Compact land use > 500.","\n")

len2 <- length(first)
cat("\nThere are",len2,"cities with English names beginning with",let1,"\n")
cities[first,1]
```

### Summaries for non-numeric data

In the `cities` data, we have three factor columns, `Type`, `Global`, 
and `Region`. The `table()` function will count the number of
observations in each category (this is called a *contingency table*):

```{r table, echo=-1}
options(width=100)
table(cities$Type)
```

This is similar to using `summary()`...

```{r summary-vs-table, echo=-1}
options(width=100)
summary(cities$Type)
```

...except that `table()` can handle combinations of categories:

```{r two-factor-table}
table(cities$Type,cities$Global)
```

> 💡You might like to try different combinations of the factors in the 
> `cities` data.

## Summaries using `apply()` and related functions

We use the `apply()` function to apply another function over whole 
columns or rows in a data frame.

Remember that we can calculate the mean for a column like this:

```{r mean2}
mean(cities$Compact)
```

The `apply()` function lets us do this calculation for several columns 
at once. Things to note about the code below:

* we restrict the data to just the numeric columns in `cities` using 
  the column indices, i.e. `cities[,2:5]`
* `MARGIN=2` means columns (`MARGIN=1` means rows)
* `FUN` identifies the function we want to apply – here we calculate 
  the `mean` of each column

```{r apply-column-means}
apply(cities[,2:5], MARGIN=2, FUN=mean)
```

> 💡 try using `apply` to calculate a different statistic for each column

There are several variations on the basic `apply()` function in R. One 
of the most useful is `tapply()`, which calculates values of a
function depending on the levels (categories) of a Factor column. For
example, to calculate the `median` industrial land area (km²) for the
different levels of city `Type`:

```{r tapply-Industry-Type, echo=-1}
options(width=100)
tapply(cities$Industry, INDEX=cities$Type, FUN=median)
```

> 💡 try using `tapply` with a different column and factor...

Remember that instead of using the `data$column` syntax, we can use
indices. Check that this code gives the same answer (we're also
omitting argument names in the `tapply()` function):

```{r tapply using indices}
tapply(cities[,5], cities[,6], median)
```

An advantage of using indices is that this form makes it easier to 
write **programming code** in R using counter variables and loops – 
as we'll see later!

🔚