---
title: ''
output:
  bookdown::html_document2:
    fig_width: 6
    fig_height: 4
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
  word_document:
    toc_depth: 2
    fig_caption: yes
---


<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r addimg function, include=FALSE}
library(png)
addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
logo <- readPNG("2015-UWA-Full-Hor-CMYK-png.png") # add our own image
```

```{r save-default-par, include=FALSE}
defpar <- par(no.readonly = TRUE)
```

```{r header-image, echo=FALSE, fig.height=3, fig.width=20, message=FALSE, warning=FALSE}
par(mar=c(0,0,0,0))
windowsFonts(uwa = windowsFont("UWA"))
plot(c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
rect(1.25, 0.9, 2, 1, col="#003087", border="#003087")
rect(1.8, 0.9, 2, 1, col="#DAAA00", border="#DAAA00")
addImg(logo, x = 1.11, y = 0.6, width = 0.2)
text(1.95,0.75,pos=2,labels="School of Agriculture and Environment", cex=3,
     family="uwa", col="#003087")
text(1.95,0.6,pos=2,labels="\u2514 Environmental Science", cex=3,
     family="uwa", col="#003087")
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

The up-to-date code to generate this document is always available 
[**here**](https://github.com/Ratey-AtUWA/Learn-R/blob/main/Why-UWAcolgrad.Rmd){target="_blank"}.

# Why a UWA colour palette generator for **R**?

Colour palettes with relatively few distinctive colours are often used in the
corporate world. Apparently they confer a consistent "look and feel" to
company-branded materials, making publications, presentations, and advertising
more memorable and the brand more recognisable.

When presenting material clearly from a scientific perspective, however,
corporate colour palettes can be problematic if they include colour combinations
that are difficult to distinguish for people having colour vision impairments
such as red-green colour-blindness. Where colour comparisons are important to
interpreting data presented graphically, the contrast between the items being
compared should be visible to as many people as possible, regardless of
colour-blindness. In addition, the print copies of some scientific journals
still reproduce figures as greyscale images, so a colour palette that maintains
as much contrast as possible when converted to greyscale is preferable. This is
achieved with:

1. a *uniform gradient*, an approximately linear relationship between greyscale value and colour palette index;
2. a wide range of initial contrast (dark to light, or *vice versa*).

The importance of a wide range in contrast when converting to greyscale is
illustrated by the four versions of the same map shown in Figure
\@ref(fig:recoloured-map) below.

```{r recoloured-map, echo=FALSE, fig.height=6.25, fig.width=10, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Four versions of the same map: (a) original colours from Thebo *et al*. (2014); (b) original map converted to grayscale; (c) recoloured map used in Rate (2022); (d) recoloured map converted to grayscale. Greyscale conversions were performed in `paint.net` (Brewster, 2023).", echo=FALSE}
logo <- readPNG("./images/recoloured-map.png") # add our own image
par(mar=c(0,0,0,0))
plot(c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
addImg(logo, x = 1.5, y = 0.5, width = 1)
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

<p>&nbsp;</p>

Fortunately for staff and students working at UWA, the standard blue and gold colours in the
[UWA brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"} 
lend themselves well to generation of themed colour palettes which are
colourblind-friendly. The development of a possible UWA colour palette with
uniform colour gradient is discussed below. I have focused on the creation of
this palette in the **R** statistical computing environment (R Core Team, 2022);
many people at UWA use **R**, but many don't, so there is some comment on use of
the UWA palette in Microsoft Office^®^ software near the end.

## 1. Existing palettes available in **R** packages

There are several existing **R** packages and functions which allow generation
of uniform-gradient, colourblind-friendly palettes.

### 1.1 RColorBrewer

The `RColorBrewer` package (Neuwirth, 2022) contains 35 in-built palettes, 27 of
which are colourblind-friendly (Figure \@ref(fig:rcolorbrewer-cbf)). One of 
these (`YlGnBu`) has end-member colors which are near the UWA colours in the 
colour wheel, but not at the exact hues.

```{r rcolorbrewer-cbf, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Colourblind-friendly palettes from the `RColorBrewer` R package. Each palette is shown with the maximum allowable number of colours.", results='hold', echo=2:3}
par(mar=c(0,4,0,0))
library(RColorBrewer)
display.brewer.all(colorblindFriendly = TRUE)
text(c(2,2,2),c(29.3,10.3,6.3),pos=2,font=2,
     labels=c("Gradational","Qualitative","Diverging"))
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

<p>&nbsp;</p>

For comparison, the `RColorBrewer` colourblind-friendly palettes viewed in 
greyscale are shown below (Figure \@ref(fig:rcolorbrewer-gray):

```{r convert-to-greyscale, message=FALSE, warning=FALSE, include=FALSE}
library(imager)
png(file="./images/rcb-col.png", width = 600, height = 600)
par(mar=c(0,4,0,0))
display.brewer.all(colorblindFriendly = TRUE)
text(c(2,2,2),c(29.3,10.3,6.3),pos=2,font=2,
     labels=c("Gradational","Qualitative","Diverging"))
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
dev.off()
rcb_col <- imager::load.image("./images/rcb-col.png")
rcb_gry <- grayscale(rcb_col)
save.image(rcb_gry, "./images/rcb-gry.png")
```

```{r rcolorbrewer-gray, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Colourblind-friendly palettes from the `RColorBrewer` R package, shown as a greyscale image created using the `imager` **R** package.", echo=FALSE}
logo <- readPNG("./images/rcb-gry.png") # add our own image
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
addImg(logo, x = 0.5, y = 0.5, width = 1)
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

```{r remove-temp-files, message=FALSE, warning=FALSE, include=FALSE}
file.remove("./images/rcb-col.png")
file.remove("./images/rcb-gry.png")
```

<p>&nbsp;</p>

### 1.2 Viridis

The `viridis` and companion `viridisLite` **R** packages (Garnier *et al*. 2021)
are specifically designed to address and minimise loss of information caused by
colourblindness. The vignette supplied with the `viridis` package provides an
excellent explanation of the need for, and functions of, the package (in
RStudio, run `help(package="viridis")`, then select <u>User guides, package
vignettes and other documentation</u>, then select
<u>viridis::intro-to-viridis</u>). The `viridis` package has eight colour 
palettes which are shown as gradients in Figure \@ref(fig:show-viridis) below 
(each palette can be used to generate fewer discrete colours as well). The
`cividis` palette, a corrected version of the original `viridis`, appears to 
contain hues most closely resembling the [UWA brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"} colours.

```{r show-viridis, fig.height=6, fig.width=8, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center',out.width='60%',fig.cap="The colour palettes available in the `viridis` and `viridisLite` R packages (Garnier et al. 2021)."}
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
library(viridis)
points(seq(0,1,l=101), rep(0.92,101),col=viridis(101),pch=15,cex=4)
text(0.5,0.92,pos=3,labels="viridis",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.8,101),col=cividis(101),pch=15,cex=4)
text(0.5,0.8,pos=3,labels="cividis",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.68,101),col=magma(101),pch=15,cex=4)
text(0.5,0.68,pos=3,labels="magma",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.56,101),col=plasma(101),pch=15,cex=4)
text(0.5,0.56,pos=3,labels="plasma",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.44,101),col=inferno(101),pch=15,cex=4)
text(0.5,0.44,pos=3,labels="inferno",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.32,101),col=mako(101),pch=15,cex=4)
text(0.5,0.32,pos=3,labels="mako",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.2,101),col=rocket(101),pch=15,cex=4)
text(0.5,0.2,pos=3,labels="rocket",cex=1.5,offset=1)
points(seq(0,1,l=101), rep(0.08,101),col=turbo(101),pch=15,cex=4)
text(0.5,0.08,pos=3,labels="turbo",cex=1.5,offset=1)
```

<p>&nbsp;</p>

### 1.3 Scico

The final **R** palette package worth investigating is `scico` (pronounced
'psycho'; Pedersen and Crameri, 2023). The `psico` package can generate 35
different palettes, some of which can be categorical; my favourites are shown in
Figure \@ref(fig:show-scico-pals) below. None of the built-in colour palettes in
the `scico` package are good analogues for the [UWA brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"} colours.

```{r show-scico-pals, echo=FALSE, fig.height=4, fig.width=10, message=FALSE, warning=FALSE, out.width="70%", fig.align='center',fig.cap="Examples of gradational palettes available in the `scico` package."}
library(scico)
scico_palette_show(palettes = c("batlow","hawaii","imola","lajolla"))
mtext("batlow",3,-1, adj=0.01,col="white",cex=1.5, family="mono", font =2)
mtext("hawaii",3,-1, adj=0.99,col="black",cex=1.5, family="mono", font =2)
mtext("imola",1,-0.1, adj=0.01,col="white",cex=1.5, family="mono", font =2)
mtext("lajolla",1,-0.1, adj=0.99,col="white",cex=1.5, family="mono", font =2)
```

<p>&nbsp;</p>

## 2. Development of a draft UWA colour palette for **R**

From the previous section we saw that several packages in **R** could generate
useful colour palettes with approximately uniform [brightness] gradients. **I
would definitely recommend** trying and using uniform-gradient,
colorblind-friendly palettes of colours from any of the packages described in
Section 1. Using these palettes appropriately would add accessibility to
presentations and written teaching materials without sacrificing aesthetic
appeal. That none of these palettes perfectly captures the UWA brand should not
be any impediment to their use!

To maintain the "look and feel" of the [UWA brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"},
however, we need a colour palette which explicitly includes the brand colours,
so the remainder of this document describes the steps in development of one such
palette. The simplest and most obvious option was to try a colour gradient with
the two core colours of the UWA brand: <span style="color:white;
background-color:#003087;">&nbsp;UWA blue&nbsp;</span> and <span
style="background-color:#DAAA00;">&nbsp;UWA gold&nbsp;</span>, bracketed by
black and white. A gradient of the desired number of colours was generated using
the `colorRampPalette()` function in
**R**. As can be seen in Figure \@ref(fig:UWA-palette-initial) below, this
produced an attractive, full-contrast, gradient which was nearly uniform.

```{r UWA-palette-initial, echo=FALSE, fig.height=9, fig.width=12, message=FALSE, warning=FALSE, out.width="80%", fig.align='center', fig.cap="Analysis of initial attempt at a UWA uniform-gradient colour palette."}
np <- 24
pal0 <- c("#000000","#003087","#DAAA00","#FFFFFF")
UWAgrad <- colorRampPalette(pal0)(np)
rgbTable <- col2rgb(UWAgrad)
palette(UWAgrad)
greyWt <- ((rgbTable["red",]*0.299) + (rgbTable["green",]*0.587) +
             (rgbTable["blue",]*0.114))/255
greyAv <- ((rgbTable["red",]/3) + (rgbTable["green",]/3) +
             (rgbTable["blue",]/3))/255

par(mar=rep(0.5,4),mfrow=c(2,1),mar=c(3,3,.5,.5),mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(0:1, 0:1, ann=F, axes = F, type="n")
points(seq(0,1,l=np), rep(0.95,np), pch=22, bg=UWAgrad, cex=6) #, col=UWAgrad
text(0.5, 0.9, labels=paste("Full colour ramp, n =",np), pos=1, cex=2)
points(seq(0,1,l=np), rep(0.7,np), pch = 22,
       bg=rgb(greyWt, greyWt, greyWt), cex = 6)
text(0.5, 0.65, labels=paste("Weighted grey ramp, n =",np), pos=1, cex=2)
points(seq(0,1,l=np), rep(0.45,np), pch = 22,
       bg=rgb(greyAv, greyAv, greyAv), cex = 6)
text(0.5, 0.4, labels=paste("Average grey ramp, n =",np), pos=1, cex=2)
points(c(0.3,0.5,0.7,0.9), rep(0.15,4), pch=22, cex=4, bg=pal0)
text(0.25,0.15,pos=2,labels="Starting colours:",cex=2,offset=1)
text(c(0.1,0.3,0.5,0.7,0.9), rep(0.15,4), pos=1, cex=1.5, offset=1,
     labels=c("hex codes:",pal0))
plot(1:np, round(greyWt,2), ylim=c(0,1), xlab="Gradient sequence number", 
     ylab="Weighted greyscale value", pch=21, cex=5, bg=UWAgrad, cex.lab=1.4, cex.axis=1.4)
mtext("Dark-light gradient",3,-1.5,cex=1.5)
abline(lm(greyWt ~ c(1:np)), lty=1, col="grey92")
abline(lm(greyWt ~ c(1:np)), lty="68")
lines(c(1,np),c(0,1),col="#FFb04080",lwd=3)
```
<p>&nbsp;</p>

The palette  in Figure \@ref(fig:UWA-palette-initial) does not have a perfectly
uniform gradient, however, so in the interests of personal perfectionism the
development process continued. (I also thought that this initial palette did not
have enough blue saturation at the 'dark end'.) By iteratively substituting some
of the core colours used to generate the palette with secondary colours from the
previous palette(s), including not starting from pure black, it proved possible
to generate more uniform palettes (Figure \@ref(figUWA-palette-iterations)):

```{r UWA-palette-iterations, echo=FALSE, fig.height=9, fig.width=12, message=FALSE, warning=FALSE, out.width="80%", fig.align='center', fig.cap="Display and analysis of subsequent attempts at uniform-gradient UWA colour palettes."}
np <- 24
sat  <-  c("#000b1f","#103f8f","#DAAA00","#E4BB5E","#ffffff")
unsat <- c("#00091A","#31517d","#ac8e4b","#E6BF68","#ffffff")
UWAgrad <- colorRampPalette(sat)(np)
rgbTable <- col2rgb(UWAgrad)
palette(UWAgrad)
greyWt <- ((rgbTable["red",]*0.299) + (rgbTable["green",]*0.587) +
             (rgbTable["blue",]*0.114))/255
greyAv <- ((rgbTable["red",]/3) + (rgbTable["green",]/3) +
             (rgbTable["blue",]/3))/255
par(mar=rep(0.5,4),mfrow=c(2,1),mar=c(3,3,.5,.5),mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(1:np, round(greyWt,2), ylim=c(-0.2,1.1), xlab="Gradient sequence number", 
     ylab="Weighted greyscale value", pch=21, cex=5, bg=UWAgrad, cex.lab=1.4, cex.axis=1.4)
mtext("Dark-light gradient: Saturated",3,-1.5,cex=1.5)
abline(lm(greyWt ~ c(1:np)), lty=1, col="grey92")
abline(lm(greyWt ~ c(1:np)), lty="68")
points(seq(1,np,l=101), rep(-0.15,101), pch=15, cex=4,
       col=colorRampPalette(sat)(101))
points(seq(10,np-1,l=length(sat)), rep(0.12,length(sat)), pch=22, cex=4, bg=sat)
text(10+(np-10)/2,0.12,pos=3,labels="Starting colours:",cex=1.5,offset=1.5)
text(seq(10,np-1,l=length(sat)), rep(0.12,length(sat)), pos=1, cex=1.2, offset=1,
     labels=sat, family="mono")
# -=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-=+=-
UWAgrad <- colorRampPalette(unsat)(np)
rgbTable <- col2rgb(UWAgrad)
palette(UWAgrad)
greyWt <- ((rgbTable["red",]*0.299) + (rgbTable["green",]*0.587) +
             (rgbTable["blue",]*0.114))/255
greyAv <- ((rgbTable["red",]/3) + (rgbTable["green",]/3) +
             (rgbTable["blue",]/3))/255
plot(1:np, round(greyWt,2), ylim=c(-0.2,1.1), xlab="Gradient sequence number", 
     ylab="Weighted greyscale value", pch=21, cex=5, bg=UWAgrad, cex.lab=1.4, cex.axis=1.4)
mtext("Dark-light gradient: Unsaturated",3,-1.5,cex=1.5)
abline(lm(greyWt ~ c(1:np)), lty=1, col="grey92")
abline(lm(greyWt ~ c(1:np)), lty="68")
points(seq(1,np,l=101), rep(-0.15,101), pch=15, cex=4,
       col=colorRampPalette(unsat)(101))
points(seq(10,np-1,l=length(unsat)), rep(0.12,length(sat)), 
       pch=22, cex=4, bg=unsat)
text(10+(np-10)/2,0.12,pos=3,labels="Starting colours:",cex=1.5,offset=1.5)
text(seq(10,np-1,l=length(unsat)), rep(0.12,length(sat)), pos=1, cex=1.2, 
     offset=1, labels=unsat, family="mono")
```

<p>&nbsp;</p>

For the palettes shown in Figure \@ref(fig:UWA-palette-iterations), the full range of contrast (zero to 100%) is not quite achieved, and the saturated version has a slightly non-uniform gradient. Both are probably suitable for most application required of them (the unsaturated version is similar to `cividis` from the `viridis` **R** package. 

If you want to use these palettes in **R**, you can download code for a function to generate them [at this link](https://github.com/Ratey-AtUWA/Learn-R/blob/main/UWAcolgrad.R){target="_blank"}. Alternatively, you can run the line of R code below:

```{r}
source("https://github.com/Ratey-AtUWA/Learn-R/raw/main/UWAcolgrad.R")
```

Explanations of the arguments of this function are given in standard format in 
[the code file itself](https://github.com/Ratey-AtUWA/Learn-R/blob/main/UWAcolgrad.R){target="_blank"}.

## 3. Using the UWA colour palette in MS Office

We all use MS Office apps, even though we might not want to admit it. MS Office
allows users to change the palette of a document (*e.g*. Word, PowerPoint,
Excel) by selecting <span style="font-family:mono;">Design | Colors |
<u>C</u>ustomize colors</span>. This opens a dialog where we can edit and save a
colour palette, entering the RGB or hex colour values available via the [UWA
brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"}
(Figure \@ref(fig:custom-mso)).

```{r custom-mso, echo=FALSE, fig.height=6.5, fig.width=12, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Editing colours in a custom palette (theme colours) in a Microsoft Office app."}
logo <- readPNG("./images/Custom-colors.png") # add our own image
par(mar=c(0,0,0,0), mfrow=c(1,2))
plot(c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
addImg(logo, x = 1.5, y = 0.5, width = 1)
logo2 <- readPNG("./images/More-colors.png") # add our own image
plot(c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
addImg(logo2, x = 1.5, y = 0.5, width = 1)
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

<p>&nbsp;</p>

```{r use-custom-mso, echo=FALSE, fig.height=5.4, fig.width=12, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Selecting a custom palette in MS Office software."}
logo <- readPNG("./images/MSOffice.png") # add our own image
par(mar=c(0,0,0,0), mfrow=c(1,1))
plot(c(0,1),type="n",axes=F,bty="n",xlab="",ylab="",xaxs="i",yaxs="i")
addImg(logo, x = 1.5, y = 0.5, width = 1)
par(mar = c(3.5,3.5,0.5,0.5), font.lab=2, mgp=c(1.6,0.3,0), tcl=0.25, 
    lend="square", ljoin="mitre")
```

<p>&nbsp;</p>

The custom palette shown also includes some of the secondary colours in the [UWA brand](https://www.uniprint.uwa.edu.au/answers/visual-identity/what-colours-do-I-need-to-use-to-comply-with-the-uwa-brand){target="_blank"}
specification.

<p>&nbsp;</p>

## References

Brewster, R. (2023). `paint.net 5.0.6` (Stable 5.6.8544.33530); dotPDN LLC,  [https://www.getpaint.net/index.html](https://www.getpaint.net/index.html){target="_blank"}

Garnier S, Ross N, Rudis R, Camargo AP, Sciaini M, Scherer C (2021). *Rvision - Colorblind-Friendly Color Maps for R*. R package version 0.6.2. (package **viridis**) [https://sjmgarnier.github.io/viridis/](https://sjmgarnier.github.io/viridis/){target="_blank"}

Neuwirth E (2022). *RColorBrewer: ColorBrewer Palettes*. R package version 1.1-3, [https://CRAN.R-project.org/package=RColorBrewer](https://CRAN.R-project.org/package=RColorBrewer){target="_blank"}.

Pedersen T, Crameri F (2023). *scico: Colour Palettes Based on the Scientific Colour-Maps*. R package version 1.4.0, [https://CRAN.R-project.org/package=scico](https://CRAN.R-project.org/package=scico){target="_blank"}

R Core Team (2022). *R: A language and environment for statistical computing*. 
R Foundation for Statistical Computing, Vienna, Austria. 
[https://www.R-project.org/](https://www.R-project.org/){target="_blank"}.

Rate, A.W. (2022). Urban Ecosystems: Soils and the Rise and Fall of Cities. **In** Rate, A.W. (Ed.), *Urban Soils: Principles and Practice*. Springer-Nature; Cham, Switzerland. [https://doi.org/10.1007/978-3-030-87316-5](https://doi.org/10.1007/978-3-030-87316-5){target="_blank"}. 

Thebo, A. L., Drechsel, P., & Lambin, E. F. (2014). Global assessment of urban and peri-urban agriculture: irrigated and rainfed croplands. *Environmental Research Letters*, **9**, 9pp. [https://doi.org/10.1088/1748-9326/9/11/114002](https://doi.org/10.1088/1748-9326/9/11/114002){target="_blank"} 

